Still Missing a Cool Title

### 2006-05-31 talk/dls2006/paper.tex
add two authors
Yay.  Only one XXX left

\maketitle

@@

However, the bytecode interpreter plays a double role, at two different
levels.  The so-called Object Spaces are \textit{domains} in the abstract
interpretation terminology.  By design, we cleanly separated these
domains from the bytecode interpreter core; the latter is only
responsible for decoding the bytecodes of an application and emulating
the corresponding stack machine.  It treats all actual application-level
objects as black boxes, and dispatches all operations on them to the
Object Space.  The Standard Object Space is a concrete domain, in which
objects are the concrete Python objects of the various built-in types:
lists, dictionaries, and so on.  By opposition, the Flow Object Space is
really an abstract domain.  It handles objects that are placeholders.
Its lattice order is shown in figure \ref{flowlattice}.

@@

\item $NullableStr$, $NullableInst(class)$ -- a string or
      \texttt{None}; resp. an instance or \texttt{None}.
\end{itemize}
%
Figures \ref{latticeoverall} and \ref{latticedetail} show how these
types are ordered to form a lattice.  We mostly use its structure of
[join-semilattice] only.

@@

Of course, this would be pointless if the annotation did not give
precise enough information for our needs.  We must describe a detail of
the abstract interpretation engine that is critical for precision: the
propagation of conditional types.  Consider the following source code
fragment:

\begin{verbatim}
    if isinstance(x, MyClass):
        f(x)
    else:
        g(x)
\end{verbatim}

Although the type of \texttt{x} may be some parent class of
\texttt{MyClass}, it can be deduced to be of the more precise type
$Inst(MyClass)$ within the positive branch of the \texttt{if}.
(Remember that our graphs are in SSI form, which means that the
\texttt{x} inside each basic block is a different Variable with a
possibly different type as annotation.) XXX flow sensivity and SSA

This is implemented by introducing an extended family of types for
boolean values:

$$
Bool(v_1: (t_1, f_1), v_2: (t_2, f_2), ...)
$$

where the $v_n$ are variables and $t_n$ and $f_n$ are types.  The
result of a check, like \texttt{isintance()} above, is typically
annotated with such an extended $Bool$.  The meaning of the type is as
follows: if the run-time value of the boolean is True, then we know
that each variable $v_n$ has a type at most as general as $t_n$; and
if the boolean is False, then each variable $v_n$ has a type at most
as general as $f_n$.  This information is propagated from the check
operation to the exit of the block via such an extended $Bool$ type,
and the conditional exit logic in the type inference engine uses it to
trim the types it propagates into the next blocks (this is where the
\textit{meet} of the lattice is used).

With the help of the above technique, we achieve a reasonable precision
in small examples.  For larger examples, a different, non-local
technique is required: the specialization of functions. XXX use 'polymorphism'

As described in the introduction, the most important downside of our
approach is that automatic specialization is a potential
performance-killer.  We \textit{do} support specialization, however: we can
generate several independently-annotated copies of the flow graphs of
certain functions.  When annotating RPython programs, such
specialization does not happen automatically: we rely on hints provided
by the programmer in the source code, in the form of flags attached to
function objects.  As we had this trade-off in mind when we wrote the
Python interpreter of PyPy, we only had to add a dozen or so hints in
the end.

>>>>

\numberofauthors{2}
\author{
\alignauthor Armin Rigo\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{Institut für Informatik}\\ 
       \affaddr{Universitätsstraße 1}\\
       \affaddr{D-40225 Düsseldorf}\\
       \affaddr{Deutschland}\\
       \email{arigo@tunes.org}
\alignauthor Samuele Pedroni\\
       \affaddr{AB Strakt}\\
       \affaddr{Norra Ågatan 10A}\\
       \affaddr{416 64  Göteborg}\\
       \affaddr{Sweden}\\
       \email{pedronis@strakt.com}
}
\date{31 May 2006}
\maketitle

@@

However, the bytecode interpreter plays a double role, at two different
levels.  The so-called Object Spaces are \textit{domains} in the abstract
interpretation terminology.  By design, we cleanly separated these
domains from the bytecode interpreter core; the latter is only
responsible for decoding the bytecodes of an application and emulating
the corresponding stack machine.  It treats all actual application-level
objects as black boxes, and dispatches all operations on them to the
Object Space.  The Standard Object Space is a concrete domain, in which
objects are the concrete Python objects of the various built-in types:
lists, dictionaries, and so on.  By opposition, the Flow Object Space is
really an abstract domain.  It handles objects that are placeholders.
Its lattice order (actually a join-semilattice only) is shown in
figure \ref{flowlattice}.

@@

\item $NullableStr$, $NullableInst(class)$ -- a string or
      \texttt{None}; resp. an instance or \texttt{None}.
\end{itemize}
%
Figures \ref{latticeoverall} and \ref{latticedetail} show how these
types are ordered to form a lattice.  We mostly use its structure of
join-semilattice only.

@@

Of course, this would be pointless if the annotation did not give
precise enough information for our needs.  We must describe a detail of
the abstract interpretation engine that is critical for precision: the
propagation of conditional types.  Consider the following source code
fragment:
%
\begin{verbatim}
    if isinstance(x, MyClass):
        f(x)
    else:
        g(x)
\end{verbatim}
%
Although the type of \texttt{x} may be some parent class of
\texttt{MyClass}, it can be deduced to be of the more precise type
$Inst(MyClass)$ within the positive branch of the \texttt{if}.
\footnote{Remember that our graphs are in SSI form, which means that the
\texttt{x} inside each basic block is a different Variable with a
possibly different type as annotation.  Given that our type inference
is not flow-sensitive, SSI gives an advantage over SSA here.}

This is implemented by introducing an extended family of types for
boolean values:
%
$$
Bool(v_1: (t_1, f_1), v_2: (t_2, f_2), ...)
$$
%
where the $v_n$ are variables and $t_n$ and $f_n$ are types.  The
result of a check, like \texttt{isintance()} above, is typically
annotated with such an extended $Bool$.  The meaning of the type is as
follows: if the run-time value of the boolean is True, then we know
that each variable $v_n$ has a type at most as general as $t_n$; and
if the boolean is False, then each variable $v_n$ has a type at most
as general as $f_n$.  This information is propagated from the check
operation to the exit of the block via such an extended $Bool$ type,
and the conditional exit logic in the type inference engine uses it to
trim the types it propagates into the next blocks (this is where the
\textit{meet} of the lattice is used).

With the help of the above technique, we achieve a reasonable precision
in small examples.  For larger examples, a different, non-local
technique is required: the specialization of type-polymorphic functions.

As described in the introduction, the most important downside of our
approach is that automatic specialization is a potential
performance-killer.  So we decided that in the normal case, all calls to
a given function contribute to a single annotation of that function.
Inside the function graph, we propagate the \textit{join} of the types
of the actual parameters of the call sites.  We \textit{do} support
specialization, however: we can generate several independently-annotated
copies of the flow graphs of certain functions.  When annotating RPython
programs, such specialization does not happen automatically: we rely on
hints provided by the programmer in the source code, in the form of
flags attached to function objects.  As we had this trade-off in mind
when we wrote the Python interpreter of PyPy, we only had to add a dozen
or so hints in the end.
